Lifestyle Choice Matters: Exploring Population Health Disparities and the Effect of Tobacco and Alcohol on Life Expectancy

### 2023-11-03 outputs/paper/paper.qmd
Upload introduction, abstract, and references

---
title: "Life expectancy"
author: "Yufei Liu"
thanks: "Code and data are available at: https://github.com/Florence-Liu/life_expectancy"
date: today
date-format: long
abstract: "This paper "
format: pdf
number-sections: true
toc: false
bibliography: references.bib
---

@@

ashdjk
asdhkja
askdhka

ahdjkajks

adkajhk

akjdha

@@

The datasets used in this paper was obtained from World Health Organization (WHO) Global Health Observatory data repository, and is publicly available from WHO website [@citeWHO]. We utilized 3 datasets: life expectancy and Health life expectancy by WHO region [@citeLife], SDG Target 3.5 Substance abuse by WHO region [@citeAlcohol], and SDG Target 3.a Tobacco control by WHO region [@citeTobacco]. The life expectancy dataset contains life expectancy and healthy life expectancy at birth and at age 60 for different sexes and WHO regions in year 2000, 2010, 2015, and 2019. 
The life expectancy values were estimated based on mortality data from civil registration. The substance abuse dataset contains total alcohol per capita (aged 15+) consumption (total APC) for different sexes and WHO regions in year 2000, 2005, 2010, 2015, and 2019. The tobacco control dataset contains the percentage of the population aged 15+ who currently use tobacco products based on population-based surveys for different sexes and WHO regions in year 2000, 2005, 2010, 2015, 2018, 2019, and 2020.

We specifically selected life expectancy at age 60 since it would be more associated with either tobacco or alcohol usage than life expectancy at birth. Then we merged the three datasets into one by region, sex, and year. Our cleaned data for analysis contains variables:

@@

```{r}
#| label: fig-alcohol
#| fig-cap: The effect of total alcohol per capita consumption on life expectanct at age 60 for different sex
#| echo: false
#| message: false
#| warning: false
data |> ggplot(aes(x=alcohol_consumption, y=life_expectancy, color=sex)) +
  geom_point() +
  #facet_grid(.~sex) +
  theme_light() +
  labs(x = "Total alcohol per capita consumption in litre",
       y = "Life expectancy at age 60 in years",
       color = "Sex") +
  scale_color_manual(values = c("indianred3","steelblue3")) +
  geom_smooth(method = "lm", se = F, linetype = "dashed") +
  geom_smooth(method = "lm", se = F, color = "darkgreen")
```
@fig-year shows life expectancy at age 60 for different years and different sexes. In general, life expectancy at age 60 increased from 2000 to 2019 for both sexes. The boxplots also shows that in each year, the mean life expectancy at age 60 for female is much higher than that for male along with larger range and variance. Also, the distribution of life expectancy for both sexes are skewed with one whisker longer than the other, indicating more observations centered at higher values. There is no outliers for all the boxplots.

@@

kable(table1, col.names = c("Year","Mean<br>life expectancy (year)", 
                            "Mean<br>prevalence of tobacco (%)",
                            "Mean<br>alcohol consumption (litre)"),
      escape = FALSE, "html")
```
@tbl-year and @tbl-region shows summary tables for the mean values of life expectancy at age 60, prevalence of tobacco, and alcohol consumption across years and regions. We could see that the mean life expectancy increased from 2000 to 2019, which is consistent with @fig-year, while the mean prevalence of tobacco decreased across years and the mean alcohol consumption is relatively stable with minor changes. The potential inverse relationship between prevalence of tobacco and life expectancy is consistent with what we found in @fig-tobacco. For different WHO regions, the three values differ a lot. Africa has the lowest mean life expectancy at age 60 while Americas has the highest mean life expectancy at age 60. However, Africa also has the lowest prevalence of tobacco, which contradicts our previous findings. Also, the mean alcohol consumption has a large range with the lowest value at 0.3 for Eastern Mediterranean and highest value at 10.2 for Europe. These findings in region difference could be further investigated since there might be potential cultural and socio-economic reasons behind.

@@

@@

Also for the alcohol per capita consumption, the model suggested that the alcohol consumption is associated with increased life expectancy at age 60. Other than tobacco products, the effect of alcohol on health reveals a complex relationship. Modest drinking has been approved to be protective against some diseases, but in most cases, alcohol could be seen as a health risk factor [@citeDrinking]. In our model, we did not specify alcohol consumption level, so it could be further investigated by creating a new dummy variable indicating different drinking levels.

According to @tbl-year and @tbl-region, the values that are not consistent with our model result indicate that there might be other factors that influence the life expectancy at age 60 for a specific year or region. For different regions, the difference in economic development, medical treatments, and environmental conditions could be potential influencing factors. However, if we introduced these factors into our model, we have to consider the problem of multicollinearity that the independent variables correlates since these socio-economic factors and environmental factors could also contribute to the prevalence of tobacco and alcohol consumption.

>>>>

---
title: "Lifestyle Choice Matters: Exploring Population Health Disparities and the Effect of Tobacco and Alcohol on Life Expectancy"
author: "Yufei Liu"
thanks: "Code and data are available at: https://github.com/Florence-Liu/life_expectancy"
date: today
date-format: long
abstract: "Life expectancy is an important measurement of population health and there are many significant influencing factors such as genetic and lifestyle. This paper uses a linear regression model to analyze life expectancy, prevalence of tobacco products, and alcohol consumption data from Global Health Observatory data repository of WHO. We find as prevalence of tobacco increases, the life expectancy at age 60 decreases while as alcohol consumption increases, the life expectancy increases. We also find regional, temporal, and sexual differences in life expectancy at age 60. It implies health disparies in several aspects and encourages the importance of health lifestyle including take less tobacco use and moderate drinking that contribute to increased life expectancy. "
format: pdf
number-sections: true
toc: false
bibliography: references.bib
---

@@

Life expectancy measures the average number of years a person could expect to live, which is the key metric to assess population health. Since the life expectancy varies across years and regions, it could be used to assess the trend and patterns of population health [@citeLife]. Several socioeconomic, genetic, lifestyle, nutritional, and environmental factors influence the life expectancy [@citeInfluence].

In this paper, we investigate how lifestyle factors influence life expectancy at age 60, and focus on two specific aspects, prevalence of tobacco products and alcohol consumption. We use `R` [@citeR] to analyze life expectancy, prevalence of tobacco, and alcohol consumption data from Global Health Observatory (GHO) data repository from World Health Organization (WHO) [@citeWHO]. We construct a multiple linear regression model in which life expectancy at age 60 for all WHO regions is explained by five variables, sex, region, year, prevalence of tobacco, and alcohol consumption. We find that female has a higher life expectancy at age 60 than male across years, and life expectancy has a large region difference as well as tobacco use and alcohol consumption. We also find as the prevalence of tobacco increases, the expected life expectancy decreases while as the alcohol consumption increases, the expected life expectancy also increases. Further work could introduce more complex models to capture non-linear relationships and give region-specific studies as well as taking into account other socioeconomic and environmental factors.

The remainder of this paper is structured as follows: @sec-data discusses the data with @sec-data_description including information about data collection and data cleaning results, and @sec-data_visualization including graphs and tables representing relationships between variables and some discussions; @sec-model introduces two linear regression models we employ and specifies parameters; @sec-result shows the estimates of fitted models and criteria for comparing two models; @sec-discussion includes discussions about model result and implications with @sec-summary making a brief summary of what the paper has done, @sec-findings showing findings based on model and data visualization, and @sec-weakness talking about the weakness and future improvement.

@@

The datasets used in this paper was obtained from World Health Organization (WHO) Global Health Observatory data repository, and is publicly available from WHO website [@citeWHO]. We utilized 3 datasets: life expectancy and Health life expectancy by WHO region [@citeLifeexp], SDG Target 3.5 Substance abuse by WHO region [@citeAlcohol], and SDG Target 3.a Tobacco control by WHO region [@citeTobacco]. The life expectancy dataset contains life expectancy and healthy life expectancy at birth and at age 60 for different sexes and WHO regions in year 2000, 2010, 2015, and 2019. 
The life expectancy values were estimated based on mortality data from civil registration. The substance abuse dataset contains total alcohol per capita (aged 15+) consumption (total APC) for different sexes and WHO regions in year 2000, 2005, 2010, 2015, and 2019. The tobacco control dataset contains the percentage of the population aged 15+ who currently use tobacco products based on population-based surveys for different sexes and WHO regions in year 2000, 2005, 2010, 2015, 2018, 2019, and 2020.

We specifically selected life expectancy at age 60 since it would be more associated with either tobacco or alcohol usage than life expectancy at birth, and data in year 2000, 2010, 2015, and 2019. Then we merged the three datasets into one by region, sex, and year. Our cleaned data for analysis contains variables:

@@

```{r}
#| label: fig-alcohol
#| fig-cap: The effect of total alcohol per capita consumption on life expectanct at age 60 for different sex
#| echo: false
#| message: false
#| warning: false
data |> ggplot(aes(x=alcohol_consumption, y=life_expectancy, color=sex)) +
  geom_point() +
  #facet_grid(.~sex) +
  theme_light() +
  labs(x = "Total alcohol per capita consumption in litre",
       y = "Life expectancy at age 60 in years",
       color = "Sex") +
  scale_color_manual(values = c("indianred3","steelblue3")) +
  geom_smooth(method = "lm", se = F, linetype = "dashed") +
  geom_smooth(method = "lm", se = F, color = "darkgreen")
```

@@

kable(table1, col.names = c("Year","Mean<br>life expectancy (year)", 
                            "Mean<br>prevalence of tobacco (%)",
                            "Mean<br>alcohol consumption (litre)"),
      escape = FALSE, "html")
```

@@

@fig-year shows life expectancy at age 60 for different years and different sexes. In general, life expectancy at age 60 increased from 2000 to 2019 for both sexes. The boxplots also shows that in each year, the mean life expectancy at age 60 for female is much higher than that for male along with larger range and variance. Also, the distribution of life expectancy for both sexes are skewed with one whisker longer than the other, indicating more observations centered at higher values. There is no outliers for all the boxplots.

@tbl-year and @tbl-region shows summary tables for the mean values of life expectancy at age 60, prevalence of tobacco, and alcohol consumption across years and regions. We could see that the mean life expectancy increased from 2000 to 2019, which is consistent with @fig-year, while the mean prevalence of tobacco decreased across years and the mean alcohol consumption is relatively stable with minor changes. The potential inverse relationship between prevalence of tobacco and life expectancy is consistent with what we found in @fig-tobacco. For different WHO regions, the three values differ a lot. Africa has the lowest mean life expectancy at age 60 while Americas has the highest mean life expectancy at age 60. However, Africa also has the lowest prevalence of tobacco, which contradicts our previous findings. Also, the mean alcohol consumption has a large range with the lowest value at 0.3 for Eastern Mediterranean and highest value at 10.2 for Europe. These findings in region difference could be further investigated since there might be potential cultural and socio-economic reasons behind.

@@

Other than tobacco products, the effect of alcohol on health reveals a complex relationship. The full model we choose suggested that the alcohol consumption is associated with increased life expectancy at age 60. However, in the reduced model, when we do not consider sex impact on alcohol consumption, it shows a negative association. Based on our justification of taking into account sex interaction with alcohol consumption, it indicates an overall positive association. Additionally, evidence-based study has found that modest drinking has been approved to be protective against some diseases, but in most cases, alcohol could be seen as a health risk factor [@citeDrinking]. In our model, we did not specify alcohol consumption level, so it could be further investigated by creating a new dummy variable indicating different drinking levels.

According to @tbl-year and @tbl-region, the values that are not consistent with our model result indicate that there might be other factors that influence the life expectancy at age 60 for a specific year or region. For different regions, the difference in economic development, medical treatments, and environmental conditions could be potential influencing factors [@citeInfluence]. However, if we introduced these factors into our model, we have to consider the problem of multicollinearity that the independent variables correlates since these socio-economic factors and environmental factors could also contribute to the prevalence of tobacco and alcohol consumption.
